Unified data structures for solving optimally a set of interrelated computational geometry problems

### 2020-01-08 eurocg20 EA/article-eurocg20-ea.tex
euroCG 2020 & extended abstract

\documentclass[a4paper,UKenglish,cleveref, autoref]{socg-lipics-v2019}
%This is a template for producing LIPIcs articles. 
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling cleveref support, use "autoref"
%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle
\graphicspath{{images/}}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}
\lstset{
	language=Java,
	breaklines,
}

\title{Unified data structures for solving optimally a set of interrelated computational geometry problems} 
\titlerunning{Data structures for solving a set of computational geometry problems}%optional, please use if title is longer than one line

\author{Vasyl Tereshchenko}{
	Faculty of computer science and cybernetics, Taras Shevchenko National University of Kyiv, Ukraine}
{vtereshch@gmail.com}{}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional
%\footnote{}
\author{Semen Chudakov}{Faculty of computer science and cybernetics, Taras Shevchenko National University of Kyiv, Ukraine}
{semen.chudakov7@gmail.com}{}{}

\authorrunning{V. Tereshchenko and S. Chudakov}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Vasyl Tereshchenko and Semen Chudakov}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[500]{Theory of computation~Computational geometry}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Algorithmic tools,   Computational geometry, Interrelated problems set, Unified algorithmic environment, Concatenable queue}%TODO mandatory; please add comma-separated list of keywords

\category{}%optional, e.g. invited paper

\relatedversion{}%optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversion{A full version of the paper is available at \url{...}.}

\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\acknowledgements{}%optional

%\nolinenumbers %uncomment to disable line numbering

%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{}
\EventNoEds{2}
\EventLongTitle{36th International Symposium on Computational Geometry (SoCG 2020)}
\EventShortTitle{SoCG 2020}
\EventAcronym{SoCG}
\EventYear{2020}
\EventDate{June  23--26, 2020}
\EventLocation{Zürich, Switzerland}
\EventLogo{socg-logo}
\SeriesVolume{}
\ArticleNo{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
	The paper is devoted to the development and  of an efficient algorithmic model for solving a set of interrelated computational geometry problems. To do this, a unified algorithmic environment with unified data structures is created, which allows to implement complex use cases efficiently with respect to computational resources. We build the environment on the basis of the ``divide and conquer'' strategy. 
	Once a convex hull is key to a set of computational geometry problems, we offer a concatenable queue data structure to maintain it. The data structure is implemented in a form of a binary tree. This allows to perform operations needed in algorithm for a set of problems in $O(\log n)$ time. Furthermore we offer a way to execute the algorithms both sequentially and in parallel.
	In the future the algorithmic environment can be improved to support other computational models with similar properties for solving problems. As an example, the Voronoi diagram or the Delaunay triangulation can be considered.
\end{abstract}

@@

	Nowadays, advanced computer simulations and visualizing of complex scientific researches as well as large scale technical projects requires to simultaneously solve a set of problems. The core of this set are problems of computational geometry and computer graphics. To solve such problems it is needed to create suitable algorithmic frameworks, that would yield accurate results in real time. Existing methods (iLastic \cite{ilastik}, IMARIS \cite{imaris}, ImageJ \cite{imagej}), that are based on a set of algorithms implementations organized in a package does not result in desirable efficiency and accuracy. It is worth noting, that there are a lot of parallel algorithms designed to solve specifically certain computational geometry problems such as in \cite{aggarwal,atallah,cole,amato,chen,berkman,goodman,akl,jaja,leeuwen,reif}. Every such algorithm requires its own computational resources and is executed independently from others. In such case some identical steps, such as preprocessing and building data structures, are performed several times.

@@

	In this article we first describe how the convex hull algorithm for a static set of points is decomposed into separate stages and incorporated into our unified algorithmic environment model (UAEM). Then we provide detailed explanation on how the concatenable queue which is used in the algorithm is implemented. Finally we make complexity analysis for the algorithm and evaluate its performance.

@@

	Consider an algorithm that, in a sorted array, for each group of identical elements, deletes all but the first and the last one (if there are more than two repetitions). The idea behind the algorithm is to use two pointers technique to delete repetitions by overwriting their place with other non-repeating element. This way, additional memory usage can be reduced to a constant value. The algorithm makes one pass through the array. At each step, a constant amount of work is performed to decide whether to delete the current item. Given this, the complexity of the above algorithm is $O(n)$.

	To perform the preprocessing described above, it is needed to:

	\begin{enumerate}
		\item
		Sort points by $y$ (if $y$ coordinates are equal, the $x$ coordinates are compared).
		\item
		Delete repetitions by $y$ coordinate using the described algorithm.
		\item
		Sort points by $x$ (if $x$ coordinates are equal, the $y$ coordinates are compared).
		\item
		Delete repetitions by $x$ coordinate using the described algorithm.
	\end{enumerate}

	In the result we get a set of points for which we can apply the recursive convex hull algorithm.

@@

	At the stage of splitting the initial problem, the set of points is divided into left and right parts of equal size. Since a array-like data structure is used to store the points, this operations can be completed in $O(1)$ using the formula below:

	\begin{equation}
		M_{i,j}=\frac{i+j}{2}
	\end{equation}
	The recursion stops when there are no more than $3$ points in the set.

	For the base case, the set of points can have $2$ or $3$ elements. In the first case the highest of two points forms the upper part of the hull, and the lower - the lower part. To consider the base case of $3$ points, we introduce an additional notion of the tangent slope given by two points on the plane. For arbitrary points $a_1, a_2$ such that $x(a_1)<x(a_2)$ slope is denoted as $\lambda$:

@@

	In \cite{overmars} an algorithm for merging two convex hulls is described. The idea of this algorithm is to maintain the hull in two concatenated queues. The hull is divided into two parts. In this article it is divided into the upper and the lower sub-hulls. Then, using the two-pointer technique and the cases described in \cite{overmars}, the proper tangent line is searched. It remains to split the two queues at the found vertices that form the tangent and to merge the remaining parts. An example of performing such a procedure is shown in Fig. \ref{fig:ch_union}.

	\begin{figure}[t]
		\centering
		\includegraphics[width=0.8\textwidth, height=0.3\textheight]{ch_union}
		\caption{Merging step of two hulls}
		\label{fig:ch_union}
	\end{figure}

	It remains to consider the corner cases that arise when performing the merging. The first of these cases is related to the ambiguity of the position of the utmost points in the described representation. The leftmost point of the left hull and the rightmost point of the right hull must belong to the upper parts of the view before finding the tangent line, because otherwise such tangent may be found incorrectly. An example of such incorrect search is given in Fig. \ref{fig:incorect_search}.

@@

	We start creating the generic interface by listing its components:

	\begin{itemize}
		\item 
		Preprocessing.
		\item 
		Splitting task into sub-tasks.
		\item 
		Solving sub-tasks.
		\item 
		Merging obtained results.
		\item 
		Checking, if a given input data is a base case for the algorithm.
		\item 
		Solving the base case.
	\end{itemize}

	Each of these components will represent a function in the future interface. Here there is a clear separation between the input type for the algorithm and the type of the result it returns. These two types should be parameters of the algorithm model. One should also pay attention to the input and output types of functions  in the interface.

	A large number of computational geometry algorithms, such as computing minimal spanning tree, the Delaunay triangulation, the Voronoi diagram and the convex hull accept the list of points. However, it would be wrong to limit the input type in the algorithm interface to a list of points. The reason for this is the fact that some algorithms can use the results of other algorithms as input data. A well known example is the construction of a Delaunay triangulation based on the Voronoi diagram maintained in a special data structure.

	Listing \ref{lst:algmodel} shows the constructed algorithm model. $IT$ denotes the input type and $OT$ - output type.

@@

@@

	As shown in \cite{overmars}, the concatenable queue is the key data structure for the algorithm described above and is therefore the basis for the UAEM. Now we will focus on how to efficiently implement it in our algorithmic environment.

	Concatenable queue is an Abstract Data Type, that supports following operations:

	\begin{itemize}
		\item
		ADD\_ELEMENT();
		\item
		REMOVE\_ELEMENT();
		\item
		GET\_MINIMUM();
		\item
		CONTAINS();
		\item
		SPLIT();
		\item
		MERGE().
	\end{itemize}

	By default the elements in a concatenable queue are kept i a certain predefined order \cite[pp..~155-157]{aho}.

	In this article the concatenable queue is implemented as a binary  $B+$ tree. Its vertices are divided into non-leaf and leaf ones. The leaf vertices contain all data kept in a tree. Every vertex has a pointer to its left and right child. For the leaf vertices those pointers point to the left and right neighboring leaf vertices or $null$ if the vertex is utmost in the tree. 
	Additionally every vertex keeps a pointer to a vertex with the largest element in its left sub-tree, which  allows to perform binary search \cite[pp..~155-157]{aho}. The height of each vertex is kept for the balancing during the split and merge operations. It is measured as a maximum amount of steps it is possible to do in order to reach a leaf vertex.

	From now we will go into details on how this data structure is implemented.

	The contains operations is pretty straightforward and uses binary search over the tree so its complexity is $O(\log n)$, where $n$ hereafter denotes the numbers of vertices in the queue.

	Algorithm of inserting an element in a queue looks like as follows. First, the position of a new vertex is searched. Then the connections between adjacent leaf vertices are broken to insert a new one. Going back a new non-leaf vertex is created - the parent for the new element and one of its neighbor. The algorithm is formally described on the Listing \ref{lst:insert-alg}.

	\begin{figure}[htbp]
		\begin{lstlisting}[caption={Queue insertion algorithm},label={lst:insert-alg},captionpos=b]
Node insert(Node node, int e):
    Node result = nil
    if e <= node.leftSubtreeMax.data:
        if node.isLeaf:
            if e == node.data:
                node.data = e
            else:
                Node createdLeaf = createLeafBetween(e, node.left, node)
                result = Node(createdLeaf, createdLeaf, node)
        else:
            node.left = insert(node.left, e)
    else:
        if node.isLeaf:
            Node createdLeaf = createLeafBetween(e, node, node.right)
            result = Node(node, node, createdLeaf)
        else:
            node.right = insert(node.right, e)

    if result == nil:
        result = node

    updateHeight(result)
    return result

		\end{lstlisting}
	\end{figure}

	Here the \textit{updateHeight} subroutine updates the height value for a given vertex, the \textit{createLeafBetween} subroutine breaks the connection between adjacent vertices ot insert a new one.

	On the first step we find out if $leftSubtreeMax$ point to an element with greater value than the value to be inserted $e$. If so, then, if current element is a leaf, a new element is created between the current element and its left neighbor. Otherwise the search proceeds on the left sub-tree of the current element. The case, when $leftSubtreeMax$ is smaller than the value $e$ is analogous. The procedure ends with updating the height on a newly created element. The element is returned as its result value.

	Since on every step we perform a constant amount of work, the complexity of the procedure is $O(h)=O(\log n)$, where $h$ hereafter denotes the height of the tree. The operation of removing an element from the queue is performed analogously.

	The split operation is a bit more complex. As an input the procedure takes current element and value based on which the split is performed. As a result two independent queues are formed. The value, by which the split has been performed, belongs to the left sub-queue. The procedure is formally described on the Listing \ref{lst:split-alg}.

	\begin{lstlisting}[caption={Queue split algorithm},label={lst:split-alg},captionpos=b]
split(Node node, int e, ConcatenableQueue leftQueue, ConcatenableQueue rightQueue):
    if node.isLeaf
        leftQueue.root = node
        leftQueue.maxNode = node
        rightQueue.minNode = node.right
        cut(node)
    else:
        if e == node.leftSubtreeMax.data:
            leftQueue.root = node.left
            leftQueue.maxNode = node.leftSubtreeMax

            rightQueue.root = node.right
            rightQueue.minNode = node.leftSubtreeMax.right

            cut(node.leftSubtreeMax)
        else if e < node.leftSubtreeMax.data:
            split(node.left, e, leftQueue, rightQueue)
            rightQueue.root = concatenateNodes(rightQueue.root, node.right, node.leftSubtreeMax)
        else:
            split(node.right, e, leftQueue, rightQueue)
            leftQueue.root = concatenateNodes(node.left,  leftQueue.root, node.leftSubtreeMax)
	\end{lstlisting}

	Here the $concatenateNodes$ procedure is used. It performs concatenation of two arbitrary elements and uses theirs heights to balance the resulting queue. The $cut$ procedure breaks connection between two adjacent leaf elements in a queue and therefore is trivial. On the first step in the $split$ operation we check, is the current element is a leaf. Is so, its connection are broken and the value of $maxNode$ is updated for the left queue as well as the value of $minNode$ for the right queue. If element is not a leaf, then the procedure continues of either left or right sub-tree. Here a special corner cases is considered, where $leftSubtreeMax$ contain the dividing value. Then the analogous action to the usual search are performed.

	The algorithm of the $concatenateNodes$ procedure is described on the Listing \ref{lst:concatenate-als}.

	\begin{lstlisting}[caption={Merging two queues},label={lst:concatenate-als},captionpos=b]
Node concatenateNodes(Node leftNode, Node rightNode, Node leftSubtreeMax) {
    if leftNode == nil:
        return rightNode
    else if rightNode == nil:
        return leftNode
    else if leftNode.height == rightNode.height:
        Node result = Node(leftSubtreeMax, leftNode, rightNode)
        updateHeight(result)
        return result
    else if leftNode.height < rightNode.height:
        rightNode.left = concatenateNodes(leftNode, rightNode.left, leftSubtreeMax)
        updateHeight(rightNode)
        return rightNode
    else:
        leftNode.right = concatenateNodes(leftNode.right, rightNode, leftSubtreeMax)
        updateHeight(leftNode)
        return leftNode
	\end{lstlisting}

	First, we consider corner cases where one of the elements is $null$. This is needed to ensure correctness of the recursion. Then, if the left element is higher than the right, one step down is taken for the left element. If the right element is higher - we take a step down for it. If the heights are equal, the joining point is found and a new element must be created. At each step, it is necessary to update the height of current element because it changes.

	We begin analyzing the complexity of the $split$ procedure by determining the complexity of the $concatenateNodes$ procedure. At each iteration, a step is performed either to the left son of the current element or to the right one. The execution of the recursive procedure finishes by merging two elements. Since each step moves us down one level and a constant amount of work is performed for each level, the total complexity of the $concatenateNodes$ procedure is $O(h)=O(\log n)$.

	The $split$ procedure uses the $concatenateNodes$ function as a subroutine. The complexity of a $split$ call is equal to the complexity of $concatenateNodes$. The number of recursive $split$ calls for one such operation is $\log n$, so the total complexity of the procedure $\log^2 n$.

	The merge operation of two queues is reduced to the clamping of their root vertices by the $concatenateNodes$ procedure, so its complexity is $O(\log n)$.

@@

		\begin{enumerate}
			\item
			Preprocessing $O(n\log n)$.
			\item
			Recursive descent and splitting the set into 2 parts $O(1)$.
			\item
			Recursive ascent and merging parts of the convex hull $O(\log n)$.
			\begin{alphaenumerate}
				\item
				Transfer of the utmost points to upper parts of convex hulls $O(\log n)$.
				\item
				Finding the tangent for the upper parts of the hulls $O(\log n)$.
				\item
				Splitting and merging the upper parts $O(\log^2 n)$.
				\item
				Moving the utmost points to the bottom of the hulls $O(\log n)$.
				\item
				Finding the tangent for the upper parts of the hulls $O(\log n)$.
				\item
				Splitting and merging the upper parts $O(\log^2 n)$.
				\item
				Merging the lower parts of the hull $O(\log n)$.
				\item
				Normalization of the obtained lower part $O(\log n)$.
			\end{alphaenumerate}
		\end{enumerate}

@@

\bibliography{article-SoCG}
\end{document}

>>>>

% for section-numbered lemmas etc., use "numberwithinsect"
\documentclass[a4paper,english,numberwithinsect,notab]{eurocg20-submission}

% the recommended bibstyle
\bibliographystyle{plainurl}
\graphicspath{{images/}}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}
\lstset{
	language=Java,
	breaklines,
}

% -------------------------------------------------------------------
\usepackage{microtype} % if unwanted, comment out

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand\polylog{{\rm polylog}}

% Author metadata::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Unified data structures for solving optimally a set of interrelated computational geometry problems}
\titlerunning{Data structures for solving a set of computational geometry problems}%optional, please use if title is longer than one line

\author[1]{Vasyl Tereshchenko}
\author[2]{Semen Chudakov}
\affil[1]{
	Faculty of computer science and cybernetics, Taras Shevchenko National University of Kyiv, Ukraine\\
  \texttt{vtereshch@gmail.com}}
\affil[2]{Faculty of computer science and cybernetics, Taras Shevchenko National University of Kyiv, Ukraine\\
  \texttt{semen.chudakov7@gmail.com}}
%mandatory. First: Use abbreviated first/middle names.
%Second (only in severe cases): Use first author plus 'et. al.'
\authorrunning{V. Tereshchenko and S. Chudakov}
\ArticleNo{100}

% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
	This paper is devoted to the development and  of an efficient algorithmic model for solving a set of interrelated computational geometry problems. To do this, a unified algorithmic environment with unified data structures is created, which allows to implement complex use cases efficiently with respect to computational resources. We build the environment on the basis of the ``divide and conquer'' strategy. 
 	Once a convex hull is key to a set of computational geometry problems, we offer a concatenable queue data structure to maintain it. The data structure is implemented in a form of a binary tree. This allows to perform operations needed in algorithm for a set of problems in $O(\log n)$ time. Furthermore we offer a way to execute the algorithms both sequentially and in parallel.
\end{abstract}

@@

	Nowadays, advanced computer simulations and visualizing of complex scientific researches as well as large scale technical projects requires to simultaneously solve a set of problems. To solve such problems it is needed to create suitable algorithmic frameworks, that would yield accurate results in real time. Existing methods (iLastic \cite{ilastik}, IMARIS \cite{imaris}, ImageJ \cite{imagej}), that are based on a set of algorithms implementations organized in a package do not result in desirable efficiency and accuracy. It is worth noting, that there are a lot of parallel algorithms designed to solve specifically certain computational geometry problems such as in \cite{aggarwal,atallah,cole,amato,chen,berkman,goodman,akl,jaja,leeuwen,reif}. Every such algorithm requires its own computational resources and is executed independently from others. In such case some identical steps, such as preprocessing and building data structures, are performed several times.

@@

	%In this article we first describe how the convex hull algorithm for a static set of points is decomposed into separate stages and incorporated into our unified algorithmic environment model (UAEM). Then we provide detailed explanation on how the concatenable queue which is used in the algorithm is implemented. Finally we make complexity analysis for the algorithm and evaluate its performance.

@@

	Consider an algorithm that, in a sorted array, for each group of identical elements, deletes all but the first and the last one. The idea behind the algorithm is to use two pointers technique to delete repetitions by overwriting their place with other non-repeating element. At each step, constant work is performed to decide whether to overwrite the current item. Given this, the complexity of the above algorithm is $O(n)$.

	%To perform the preprocessing described above, it is needed to:

	%\begin{enumerate}
	%	\item
	%	Sort points by $y$ (if $y$ coordinates are equal, the $x$ coordinates are compared).
	%	\item
	%	Delete repetitions by $y$ coordinate using the described algorithm.
	%	\item
	%	Sort points by $x$ (if $x$ coordinates are equal, the $y$ coordinates are compared).
	%	\item
	%	Delete repetitions by $x$ coordinate using the described algorithm.
	%\end{enumerate}

	%Finally, we get a set of points for which we can apply the recursive convex hull algorithm.

@@

	At the stage of splitting the initial problem, the set of points is divided into left and right parts of equal size. Since a array-like data structure is used to store the points, this operations can be done in $O(1)$ by computing middle positiong in the array.

	%\begin{equation}
	%	M_{i,j}=\frac{i+j}{2}
	%\end{equation}
	The recursion stops when there are $2$ or $3$ elements in the array. First case is triial. To consider the base case of $3$ points, we introduce a notion of the tangent slope given by two points on the plane. For arbitrary points $a_1, a_2$ such that $x(a_1)<x(a_2)$ slope is denoted as $\lambda$:

@@

	%In \cite{overmars} an algorithm for merging two convex hulls is described. The idea of this algorithm is to maintain the hull in two concatenated queues. The hull is divided into two parts. In this article it is divided into the upper and the lower sub-hulls. Then, using the two-pointer technique and the cases described in \cite{overmars}, the proper tangent line is searched. It remains to split the two queues at the found vertices that form the tangent and to merge the remaining parts. An example of performing such a procedure is shown in Fig. \ref{fig:ch_union}.

	%\begin{figure}[t]
	%	\centering
	%	\includegraphics[width=0.8\textwidth, height=0.3\textheight]{ch_union}
	%	\caption{Merging step of two hulls}
	%	\label{fig:ch_union}
	%\end{figure}

	In \cite{overmars} an algorithm for merging two convex hulls is described. It remains to consider the corner cases that arise when performing the merging. The first of these cases is related to the ambiguity of the position of the utmost points in the described representation. The leftmost point of the left hull and the rightmost point of the right hull must belong to the upper parts of the view before finding the tangent line, because otherwise such tangent may be found incorrectly. An example of such incorrect search is given in Fig. \ref{fig:incorect_search}.

@@

	The proposed generic interface for the ``divide-and-conquer'' algorithm is described on the Listing \ref{lst:algmodel}.

@@

	%We start creating the generic interface by listing its components:

	%\begin{itemize}
	%	\item 
	%	Preprocessing.
	%	\item 
	%	Splitting task into sub-tasks.
	%	\item 
	%	Solving sub-tasks.
	%	\item 
	%	Merging obtained results.
	%	\item 
	%	Checking, if a given input data is a base case for the algorithm.
	%	\item 
	%	Solving the base case.
	%\end{itemize}

	%Each of these components will represent a function in the future interface. Here there is a clear separation between the input type for the algorithm and the type of the result it returns. These two types should be parameters of the algorithm model. One should also pay attention to the input and output types of functions  in the interface.

	%A large number of computational geometry algorithms, such as computing minimal spanning tree, the Delaunay triangulation, the Voronoi diagram and the convex hull accept the list of points. However, it would be wrong to limit the input type in the algorithm interface to a list of points. The reason for this is the fact that some algorithms can use the results of other algorithms as input data. A well known example is the construction of a Delaunay triangulation based on the Voronoi diagram maintained in a special data structure.

	%Listing \ref{lst:algmodel} shows the constructed algorithm model. $IT$ denotes the input type and $OT$ - output type.

%	\begin{lstlisting}[caption={Algorithm model based on the ``divide-and-conquer'' principle},label={lst:algmodel},captionpos=t,float,abovecaptionskip=-\medskipamount]
%interface DaCAlgorithm[IT, OT]:
%    boolean isBaseCase(IT input)
%    int inputSize(IT input)

%    OT solveBaseCase(IT input)
%    IT preprocess(IT input)

%    OT merge(OT first, OT second)
%    Pair[IT, IT] divide(IT input)
%	\end{lstlisting}

@@

	As shown in \cite{overmars}, the concatenable queue is the key data structure for the algorithm described above and is therefore the basis for the UAEM.

	Concatenable queue is an Abstract Data Type. Its operations are shown on Table \ref{table:cqueue-operations} together with theirs time complexities in our implementation:

	\begin{table}[htbp]
		\caption{Concatebable queue operations}
		\label{table:cqueue-operations}
		\begin{center}
			\begin{tabular}{|c|c|}
				\hline
				\textbf{Operation} & \textbf{Complexity} \\
				\hline
				ADD\_ELEMENT() & $O(\log n)$ \\
				\hline
				REMOVE\_ELEMENT() & $O(\log n)$ \\
				\hline
				GET\_MINIMUM() & $O(\log n)$ \\
				\hline
				CONTAINS() & $O(\log n)$ \\
				\hline
				SPLIT() & $O(\log^2 n)$ \\
				\hline
				MERGE() & $O(\log n)$ \\
				\hline
			\end{tabular} 
		\end{center}
	\end{table}

%	\begin{itemize}
%		\item
%		ADD\_ELEMENT();
%		\item
%		REMOVE\_ELEMENT();
%		\item
%		GET\_MINIMUM();
%		\item
%		CONTAINS();
%		\item
%		SPLIT();
%		\item
%		MERGE().
%	\end{itemize}

	%By default the elements in a concatenable queue are kept i a certain predefined order \cite[pp..~155-157]{aho}.

	%In this article the concatenable queue is implemented as a binary  $B+$ tree. Its vertices are divided into non-leaf and leaf ones. The leaf vertices contain all data kept in a tree. Every vertex has a pointer to its left and right child. For the leaf vertices those pointers point to the left and right neighboring leaf vertices or $null$ if the vertex is utmost in the tree. 
	%Additionally every vertex keeps a pointer to a vertex with the largest element in its left sub-tree, which  allows to perform binary search \cite[pp..~155-157]{aho}. The height of each vertex is kept for the balancing during the split and merge operations. It is measured as a maximum amount of steps it is possible to do in order to reach a leaf vertex.

	%From now we will go into details on how this data structure is implemented.

	%The contains operations is pretty straightforward and uses binary search over the tree so its complexity is $O(\log n)$, where $n$ hereafter denotes the numbers of vertices in the queue.

	%Algorithm of inserting an element in a queue looks like as follows. First, the position of a new vertex is searched. Then the connections between adjacent leaf vertices are broken to insert a new one. Going back a new non-leaf vertex is created - the parent for the new element and one of its neighbor. The algorithm is formally described on the Listing \ref{lst:insert-alg}.

	%The algorithm of inserting an element in a queue is described on the Listing \ref{lst:insert-alg}.

%	\begin{figure}[htbp]
%		\begin{lstlisting}[caption={Queue insertion algorithm},label={lst:insert-alg},captionpos=b]
%Node insert(Node node, int e):
%  Node result = nil
%    if e <= node.leftSubtreeMax.data:
%        if node.isLeaf:
%            if e == node.data:
%                node.data = e
%            else:
%                Node createdLeaf = createLeafBetween(e, node.left, node)
%                result = Node(createdLeaf, createdLeaf, node)
%        else:
%            node.left = insert(node.left, e)
%    else:
%        if node.isLeaf:
%            Node createdLeaf = createLeafBetween(e, node, node.right)
%            result = Node(node, node, createdLeaf)
%        else:
%            node.right = insert(node.right, e)
%
%    if result == nil:
%        result = node
%
%    updateHeight(result)
%    return result
%
%		\end{lstlisting}
%	\end{figure}

%	Here the \textit{updateHeight} subroutine updates the height value for a given vertex, the \textit{createLeafBetween} subroutine breaks the connection between adjacent vertices ot insert a new one.

%	On the first step we find out if $leftSubtreeMax$ point to an element with greater value than the value to be inserted $e$. If so, then, if current element is a leaf, a new element is created between the current element and its left neighbor. Otherwise the search proceeds on the left sub-tree of the current element. The case, when $leftSubtreeMax$ is smaller than the value $e$ is analogous. The procedure ends with updating the height on a newly created element. The element is returned as its result value.

%	Since on every recursion step we perform a constant amount of work, the complexity of the procedure is $O(h)=O(\log n)$, where $h$ hereafter denotes the height of the tree. The operation of removing an element from the queue is performed analogously.

%	The split operation is a bit more complex. 
%	As an input the procedure takes current element and value based on which the split is performed. As a result two independent queues are formed. The value, by which the split has been performed, belongs to the left sub-queue. 
%	Its algorithm is described on the Listing \ref{lst:split-alg}.

%	\begin{lstlisting}[caption={Queue split algorithm},label={lst:split-alg},captionpos=b]
%split(Node node, int e, ConcatenableQueue leftQueue, ConcatenableQueue rightQueue):
%    if node.isLeaf
%        leftQueue.root = node
%        leftQueue.maxNode = node
%        rightQueue.minNode = node.right
%        cut(node)
%    else:
%        if e == node.leftSubtreeMax.data:
%            leftQueue.root = node.left
%            leftQueue.maxNode = node.leftSubtreeMax
%
%            rightQueue.root = node.right
%            rightQueue.minNode = node.leftSubtreeMax.right
%
%            cut(node.leftSubtreeMax)
%        else if e < node.leftSubtreeMax.data:
%            split(node.left, e, leftQueue, rightQueue)
%            rightQueue.root = concatenateNodes(rightQueue.root, node.right, node.leftSubtreeMax)
%        else:
%            split(node.right, e, leftQueue, rightQueue)
%            leftQueue.root = concatenateNodes(node.left,  leftQueue.root, node.leftSubtreeMax)
%	\end{lstlisting}

%	Here the $concatenateNodes$ procedure is used. It performs concatenation of two arbitrary elements and uses theirs heights to balance the resulting queue. The $cut$ procedure breaks connection between two adjacent leaf elements in a queue and therefore is trivial. 
%	On the first step in the $split$ operation we check, is the current element is a leaf. Is so, its connection are broken and the value of $maxNode$ is updated for the left queue as well as the value of $minNode$ for the right queue. If element is not a leaf, then the procedure continues of either left or right sub-tree. Here a special corner cases is considered, where $leftSubtreeMax$ contain the dividing value. Then the analogous action to the usual search are performed.

%	The algorithm of the $concatenateNodes$ procedure is described on the Listing \ref{lst:concatenate-als}.

%	\begin{lstlisting}[caption={Merging two queues},label={lst:concatenate-als},captionpos=b]
%Node concatenateNodes(Node leftNode, Node rightNode, Node leftSubtreeMax) {
%    if leftNode == nil:
%        return rightNode
%    else if rightNode == nil:
%        return leftNode
%    else if leftNode.height == rightNode.height:
%        Node result = Node(leftSubtreeMax, leftNode, rightNode)
%        updateHeight(result)
%        return result
%    else if leftNode.height < rightNode.height:
%        rightNode.left = concatenateNodes(leftNode, rightNode.left, leftSubtreeMax)
%        updateHeight(rightNode)
%        return rightNode
%    else:
%        leftNode.right = concatenateNodes(leftNode.right, rightNode, leftSubtreeMax)
%        updateHeight(leftNode)
%        return leftNode
%	\end{lstlisting}

%	First, we consider corner cases where one of the elements is $null$. This is needed to ensure correctness of the recursion. Then, if the left element is higher than the right, one step down is taken for the left element. If the right element is higher - we take a step down for it. If the heights are equal, the joining point is found and a new element must be created. At each step, it is necessary to update the height of current element because it changes.

%	We begin analyzing the complexity of the $split$ procedure by determining the complexity of the $concatenateNodes$ procedure. At each iteration, a step is performed either to the left son of the current element or to the right one. The execution of the recursive procedure finishes by merging two elements. Since each step moves us down one level and a constant amount of work is performed for each level, the total complexity of the $concatenateNodes$ procedure is $O(h)=O(\log n)$.

%	The $split$ procedure uses the $concatenateNodes$ function as a subroutine. The complexity of a $split$ call is equal to the complexity of $concatenateNodes$. The number of recursive $split$ calls for one such operation is $\log n$, so the total complexity of the procedure $\log^2 n$.

%	The merge operation of two queues is reduced to the clamping of their root vertices by the $concatenateNodes$ procedure, so its complexity is $O(\log n)$.

@@

		\begin{enumerate}
			\item
			Preprocessing $O(n\log n)$.
			\item
			Recursive descent and splitting the set into 2 parts $O(1)$.
			\item
			Recursive ascent and merging parts of the convex hull $O(\log n)$.
			\begin{enumerate}
				\item
				Transfer of the utmost points to upper parts of convex hulls $O(\log n)$.
				\item
				Finding the tangent for the upper parts of the hulls $O(\log n)$.
				\item
				Splitting and merging the upper parts $O(\log^2 n)$.
				\item
				Moving the utmost points to the bottom of the hulls $O(\log n)$.
				\item
				Finding the tangent for the upper parts of the hulls $O(\log n)$.
				\item
				Splitting and merging the upper parts $O(\log^2 n)$.
				\item
				Merging the lower parts of the hull $O(\log n)$.
				\item
				Normalization of the obtained lower part $O(\log n)$.
			\end{enumerate}
		\end{enumerate}

@@

\bibliography{article-eurocg20-ea}
\end{document}
